A significance test in forward stepwise regression

### 2014-03-17 paper/paper.tex
Code: finished? general projection
Added HIV code and section

Denote $y \in \real^n$ for $n$ i.i.d. measurements of the outcome variable.
Let an integer $G \geq 2$ be the number of groups of explanatory variables.
For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables or columns in group $g$. When a group
encodes a categorical variable as binary indicators for the levels of
that variable, we use the full encoding with a column for every level.
Although this introduces collinearity, our method does not require the
design matrix to have full rank.

@@

We now describe how to calculate our p-value, following the discussion of
group lasso in \cite{tests:adaptive}. For the rest of this section let
$g = \gstar$ be the index of
the group attaining the maximum on line 3 of Algorithm~\ref{algo:fs}.
%\lambda = \norm{X_g^Ty}_2/w_g$ be the value of the maximum, and
%$r = \text{rank}(X_g)$ be the rank of the maximizing group.
For a vector $u \in \real^p$, let $u_h$ denote the coordinates
of $u$ corresponding to the columns of group $h$ in the design matrix $X$.
We can rearrange the columns of $X$ to group these adjacently, so that
\[
u^TX^T = \begin{pmatrix} u_1^T X_1^T & u_2^T X_2^T & \cdots & u_G^TX_G^T \end{pmatrix}
\]
One step of the calculation will be to find an orthonormal basis for the
linear space $V = \{ u \in \real^p : u_g^T X_g^T y = 0, u_h = 0
\text{ for all } h \neq g \}$ so that we can project orthogonally to
this space. If $X_g$ is a single column and $X_g^Ty \neq 0$, then the
space is trivial and the desired orthogonal projection is the identity.
Otherwise, if $X_g$ has $p_g > 1$ columns,

Define $U_g = X_g^Ty/(w_g\norm{X_g^Ty}_2)$.
\begin{algorithm}
  \caption{Computing p-value}
  \label{algo:pval}
  \begin{algorithmic}[1]
    \REQUIRE Response $y$, design matrix $X$, index $g$ of next group to enter
    \ENSURE P-value for testing if $g$ and all other remaining groups are noise
    \STATE $\lambda \gets \norm{X_g^Ty}_2/w_g$, $r \gets \text{rank}(X_g)$
    \IF{$X_g$ is a single column}
    \STATE $P \gets 0$ is a $p \times p$ matrix
    \STATE $\sigma^2_g \gets X_g^T \Sigma X_g$
    \ELSE
    \STATE Do that
    \ENDIF
    \RETURN $A$
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Real data example}
\todo Code simulation with HIVdb data \\
\todo Write section \\

>>>>

Denote $y \in \real^n$ for $n$ i.i.d. measurements of the outcome variable.
Let an integer $G \geq 2$ be the number of groups of explanatory variables.
For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables or columns in group $g$.
% Do we need to assume $p_g \ll n$ or $\text{rank}(X_g) = p_g$?
When a group encodes a categorical variable as binary indicators for the
levels of that variable, we use the full encoding with a column for every
level. Although this introduces collinearity, our method does not require
the design matrix to have full rank.

@@

We now describe how to calculate our p-value, following the discussion of
group lasso in \cite{tests:adaptive}. For the rest of this section let
$g = \gstar$ be the index of
the group attaining the maximum on line 3 of Algorithm~\ref{algo:fs}.
%\lambda = \norm{X_g^Ty}_2/w_g$ be the value of the maximum, and
%$r = \text{rank}(X_g)$ be the rank of the maximizing group.
For a vector $u \in \real^p$, let $u_h$ denote the coordinates
of $u$ corresponding to the columns of group $h$ in the design matrix $X$.
We can rearrange the columns of $X$ to group these adjacently, so that
\[
u^TX^T = \begin{pmatrix} u_1^T X_1^T & u_2^T X_2^T & \cdots & u_G^TX_G^T \end{pmatrix}
\]
One step of the calculation will be to find an orthonormal basis for the
linear space $\vecsp_g = \{ u \in \real^p : u_g^T X_g^T y = 0, u_h = 0
\text{ for all } h \neq g \}$ so that we can project orthogonally to
this space. If $X_g$ is a single column and $X_g^Ty \neq 0$ (which
should be the case since $g$ maximizes the absolute value of this
quantity), then the
space is trivial and the desired orthogonal projection is the identity.
Otherwise, if $X_g$ has $p_g > 1$ columns the space $\vecsp_g$ will
generally have dimension $p_g-1$.
% have dimension $r-1$ for $r = \text{rank}(X_g)$. This holds except in
% degenerate cases such as $X_g = 0$ or $r \ll p_g$, but these cases are not
% likely to occur since $\norm{X_g^T y}_2/w_g$ is supposed to be maximized.
% We proceed by forming the $n \times p_g$ matrix
% $X_g - X_g X_g^T y y^T X_g/\norm{X_g^Ty}_2^2$ and performing Gram-Schmidt
% on the first $r-1$ columns. For precise theorems showing that this will
% generally work, see \cite{outerp:indep}.
We compute an orthonormal basis by Gram-Schmidt and form a
$p \times p_g$ matrix, which we denote $V_g$, by appending 0's in the
coordinates corresponding to all groups $h \neq g$ and to one column
in group $g$ (since Gram-Schmidt only produces $p_g-1$ columns).
With this we are ready to define the projection

\begin{equation}
  \begin{aligned}
    \label{eq:proj}
    P_g &= \Sigma XV_g (V_g^T X^T \Sigma X V_g)^\dagger V_g^TX^T \\
   \end{aligned}
\end{equation}
and the conditional variance

\begin{equation}
  \begin{aligned}
    \label{eq:cvar}
    \sigma^2_g &= y^TX_gX_g^T(I-P_g)\Sigma X_gX_g^Ty / (w_g^2 \norm{X_g^Ty}_2^2) \\
  \end{aligned}
\end{equation}
These simplify when $X_g$ is a single column, in which case $P_g = 0$
and $\sigma^2_g = X_g^T\Sigma X_g^T$.

Define $U_g = X_g^Ty/(w_g\norm{X_g^Ty}_2)$.
\begin{algorithm}
  \caption{Computing p-value}
  \label{algo:pval}
  \begin{algorithmic}[1]
    \REQUIRE Response $y$, design matrix $X$, index $g$ of next group to enter
    \ENSURE P-value for testing if $g$ and all other remaining groups are noise
    \STATE $\lambda \gets \norm{X_g^Ty}_2/w_g$, $r \gets \text{rank}(X_g)$
%    \IF{$X_g$ is a single column}
%    \STATE $P_g \gets 0$ is an $n \times n$ matrix
%    \STATE $\sigma^2_g \gets X_g^T \Sigma X_g$
%    \ELSE
    \STATE Compute $P_g$, $\sigma^2_g$ from equations \ref{eq:proj} and
      \ref{eq:cvar}
%    \ENDIF
    \RETURN $A$
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Real data example}
\todo Sanity check my analysis \\
\todo For rare mutations try binary wildtype/non-wt? \\
\todo Finish writing section \\

\cite{HIV} use genomic information to predict efficacy of antiretroviral
drugs in treating HIV. Quantitative measurements of drug
response/resistance were regressed on categorical covariates encoded
the presence and type of certain genetic mutations. We attempt a similar
analysis using forward stepwise, and report our P-value at each step.
Categorical covariates are encoding as groups of binary variables using
the full encoding, and weights are set to the square-root of the
number of levels for that covariate. We perform forward stepwise once
for each drug response, restricting to the subset of the data with
no missing values and requiring all categorical variables to have 2 or
greater observations in each level.

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{../figs/HIV_PI.pdf}
\caption{\small \it Results from PI dataset}
\label{fig:HIVPI}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{../figs/HIV_NRTI.pdf}
\caption{\small \it Results from NRTI dataset}
\label{fig:HIVNRTI}
\end{center}
\end{figure}

Results from two data sets are displayed in Figure~\ref{fig:HIVPI} and
Figure~\ref{fig:HIVNRTI}. Each panel shows results for a different drug,
with p-values plotted by step in forward stepwise. The solid black point
is the last p-value to fall under the 0.1 threshold and points after it
are colored red. If the \textit{last} stopping rule were used then the
p-values for including variables are the black points, and the resulting
models are shown in Tables 4 and 5. (\todo Fix this ref after re-running)

%\begin{figure}
%\begin{center}
\input{../tables/HIV_PI.tex}
%\label{tab:HIVPI}
%\end{center}
%\end{figure}

%\begin{figure}
%\begin{center}
\input{../tables/HIV_NRTI.tex}
%\label{tab:HIVNRTI}
%\end{center}
%\end{figure}
